MS-A1 Formulae and Equations

### 2021-01-26 ms-a1_formulae-and-equations.rmd
Added MS-A1_e, MS-A1_f and MS-A1_g. Added calc font, added formula box

* review substitution of numerical values into linear and non-linear algebraic expressions and equations
  + review evaluating the subject of a formula, given the value of other pronumerals in the formula
  + change the subject of a formula
  + solve problems involving formulae, including calculating distance, speed and time (with change of units of measurement as required) or calculating stopping distances of vehicles using a suitable formula
* develop and solve linear equations, including those derived from substituting values into a formula, or those developed from a word description
* calculate and interpret blood alcohol content (BAC) based on drink consumption and body weight
  + use formulae, both in word form and algebraic form, to calculate an estimate for blood alcohol content $(BAC)$, including $BAC_{Male}=\frac{10N-7.5H}{6.8M}$ and $BAC_{Female}=\frac{10N-7.5H}{5.5M}$ where $N$ is the number of standard drinks consumed, $H$ is the number of hours of drinking, and $M$ is the person's weight in kilograms.
  + determine the number of hours required for a person to stop consuming alcohol in order to reach zero BAC, eg using the formula $time=\frac{BAC}{0.015}$
  + describe the limitations of methods estimating BAC
* calculate required medication dosages for children and adults from packets, given age or weight, using Fried’s, Young’s or Clark’s formula as appropriate
  + Fried's formula: Dosage for children 1-2 years $=\frac{\text{age (in months)}\ \times\ \text{adult dosage}}{150}$
  + Young's formula: Dosage for children 1-12 years $=\frac{\text{age of child (in years)}\ \times\ \text{adult dosage}}{\text{age od child (in years)}\ +\ 150}$
  + Clark's formula: Dosage $=\frac{\text{weight in kg}\ \times\ \text{adult dosage}}{70}$

@@

>>>>

* review substitution of numerical values into linear and non-linear algebraic expressions and equations
  + review evaluating the subject of a formula, given the value of other pronumerals in the formula
  + change the subject of a formula
  + solve problems involving formulae, including calculating distance, speed and time (with change of units of measurement as required) or calculating stopping distances of vehicles using a suitable formula
* develop and solve linear equations, including those derived from substituting values into a formula, or those developed from a word description
* calculate and interpret blood alcohol content (BAC) based on drink consumption and body weight
  + use formulae, both in word form and algebraic form, to calculate an estimate for blood alcohol content $(BAC)$, including $BAC_{Male}=\frac{10N-7.5H}{6.8M}$ and $BAC_{Female}=\frac{10N-7.5H}{5.5M}$ where $N$ is the number of standard drinks consumed, $H$ is the number of hours of drinking, and $M$ is the person's weight in kilograms.
  + determine the number of hours required for a person to stop consuming alcohol in order to reach zero BAC, eg using the formula $time=\frac{BAC}{0.015}$
  + describe the limitations of methods estimating BAC
* calculate required medication dosages for children and adults from packets, given age or weight, using Fried’s, Young’s or Clark’s formula as appropriate
  + Fried's formula: Dosage for children 1-2 years $=\frac{\text{age (in months)}\ \times\ \text{adult dosage}}{150}$
  + Young's formula: Dosage for children 1-12 years $=\frac{\text{age of child (in years)}\ \times\ \text{adult dosage}}{\text{age of child (in years)}\ +\ 150}$
  + Clark's formula: Dosage $=\frac{\text{weight in kg}\ \times\ \text{adult dosage}}{70}$

@@

# Speed, Distance and Time [![](pres.svg){width=16}](https://claytonpollard.github.io/stage6_standard_presentations/MS-A1/e_speed-distance-time.pdf){target="blank"} {.tabset .tabset-fade .tabset-pills}

* review substitution of numerical values into linear and non-linear algebraic expressions and equations
  + solve problems involving formulae, including calculating distance, speed and time (with change of units of measurement as required) or calculating stopping distances of vehicles using a suitable formula

### Speed

**Definition**

**Speed** is a comparison between *distance travelled* and *time taken*

In most real world situations, an objects speed is not constant. Therefore we often use the **average speed**.

::: {.formula}

$$s=\frac{d}{t}$$

Where:

* $s$ - average speed
* $d$ - distance travelled
* $t$ - time taken

The formula is **not** on the reference sheet.

Find the average speed of a car which travels 110 km in 2 hours.

Find the average speed of a car which travels 110 km in 2 hours

$\begin{aligned}s&=\frac{d}{t}\\&=\frac{110\text{ km}}{2\text{ h}}\\&= 55\text{ km/h}\end{aligned}$

### Rearranging

The formula for speed can be rewritten to make *distance* or *time* the subject.

$$s=\frac{d}{t}$$

$$d=st$$

$$t=\frac{d}{s}$$

Jonah rides his motorcycle on a highway at an average speed of 90 km/h.

a) How far can Jonah travel in $1\frac{1}{2}$ hours?
b) How long will he take to travel 210 km? Answer in hours and minutes.

Jonah rides his motorcycle on a highway at an average speed of 90 km/h.

a) How far can Jonah travel in $1\frac{1}{2}$ hours?
b) How long will he take to travel 210 km? Answer in hours and minutes.

<div class="row">
  <div class="column">
  $\begin{aligned}d&=st\\d&=(90)\underset{\color{red}{(*)}}{\left(\color{red}{1\frac{1}{2}}\right)}\\d&=135\text{ km}\end{aligned}$
  </div>
  <div class="column">
  $(*)=$ <span style="font-family: casio">qa1$1R2</span>
  </div>
</div>

Jonah rides his motorcycle on a highway at an average speed of 90 km/h.

a) How far can Jonah travel in $1\frac{1}{2}$ hours?
b) How long will he take to travel 210 km? Answer in hours and minutes.

<div class="row">
  <div class="column">
  $\begin{aligned}t&=\frac{d}{s}\\t&=\frac{210}{90}\\t&=2.33...\text{ h}\\&\quad\color{red}{(*)}\\t&=2^\circ 20' 0''\end{aligned}$

  2 hours and 20 minutes
  </div>
  <div class="column">
  $(*)=$ <span style="font-family: casio">x</span>
  </div>
</div>

### Stopping distance

**Definition - Stopping distance**

The distance a vehicle will travel in order to come to a complete stop.

<center>stopping distance = reaction distance + braking distance</center>

You **are not** expected to memorise any formulae.

You **are** expected to be able to substitute into a given formula.

Claire is driving on a motorway at a speed of 110 kilometres per hour and has to brake suddenly. She has a reaction time of 2 seconds and a braking distance of 59.2 metres.

What is Claire's stopping distance?

Claire is driving on a motorway at a speed of 110 kilometres per hour and has to brake suddenly. She has a reaction time of 2 seconds and a braking distance of 59.2 metres.

What is Claire's stopping distance?

$\begin{aligned}110\text{ km/h}&=110000\text{ m/hr}\\&=\frac{110000}{60\times60}\text{ m/s}\\&=30.555...\text{ m/s}\end{aligned}$

Let $d$ be Claire's stopping distance.

$\begin{aligned}d&=\underbrace{\text{reaction time distance}}_{d=st}+\text{braking distance}\\&=st+\text{braking distance}\\&=(30.555...)(2)+(59.2)\\&=120.311...\\d&=120.3 \text{ m (1 d.p.)}\end{aligned}$

Claire's stopping distance is 120.3 metres.

Max was driving 60 km/h and has a reaction time of 0.8 s. Calculate the stopping distance correct to the nearest metre given the formula $\quad d=\dfrac{5vt}{18}+\dfrac{v^2}{170}\quad$ where $d$ is the stopping distance (m), $v$ is the speed (m/s) and $t$ is reaction time (s).

Max was driving 60 km/h and has a reaction time of 0.8 s. Calculate the stopping distance correct to the nearest metre given the formula $\quad d=\dfrac{5vt}{18}+\dfrac{v^2}{170}\quad$ where $d$ is the stopping distance (m), $v$ is the speed (m/s) and $t$ is reaction time (s).

\begin{aligned}d&=\frac{5vt}{18}+\frac{v^2}{170}\\d&=\frac{5(60)(0.8)}{18}+\frac{60^2}{170}\\d&=34.509...\\d&\approx35\text{ m}\end{aligned}

Max's stopping distance is 35 metres.

# Blood Alcohol Content [![](pres.svg){width=16}](https://claytonpollard.github.io/stage6_standard_presentations/MS-A1/f_blood-alcohol-content.pdf){target="blank"} {.tabset .tabset-fade .tabset-pills}

* calculate and interpret blood alcohol content (BAC) based on drink consumption and body weight
  + use formulae, both in word form and algebraic form, to calculate an estimate for blood alcohol content $(BAC)$, including $BAC_{Male}=\frac{10N-7.5H}{6.8M}$ and $BAC_{Female}=\frac{10N-7.5H}{5.5M}$ where $N$ is the number of standard drinks consumed, $H$ is the number of hours of drinking, and $M$ is the person's weight in kilograms.
  + determine the number of hours required for a person to stop consuming alcohol in order to reach zero BAC, eg using the formula $time=\frac{BAC}{0.015}$
  + describe the limitations of methods estimating BAC

### Blood Alcohol Content

**Definition**

Blood alcohol content (BAC) is a measure of the amount of alcohol in your blood.

BAC tells you the number of grams of alcohol per 100 mL of blood

For example, a BAC of 0.10 means there is 0.10 g of alcohol for every 100 mL of blood, or 1 g for every 1 L of blood.

### Estimating BAC

BAC can be estimated using the following formulae:

::: {.formula}

$$BAC_{\text{Male}}=\frac{10N-7.5H}{6.8M}\quad\text{or}\quad BAC_{\text{Female}}=\frac{10N-7.5H}{5.5M}$$

Where:
* $BAC$ - Blood alcohol content
* $N$ - Number of standard drinks consumed
* $H$ - Hours drinking
* $M$ - Mass in kilograms

These formulae are **not** on the reference sheet.

You are **not** expected to remember these formulae.

### Limitations

These formulae are only good for calculating *estimates* of a person BAC.

The formulae for estimating BAC only takes into account the number of standard drinks consumed, sex, the amount of time, and mass.

There are other factors that influence BAC including:

* Fitness
* Health
* Liver function
* Food in the stomach
* Medications

Osman is 87 kg and has consumed 3 standard drinks in the past hour. Estimate Osman's BAC to 3 decimak places using the following formula where $N$ represents the number of standard drinks, $H$ is the number of hours drinking and $M$ is his mass in kilograms.

$$BAC_{\text{Male}}=\frac{10N-7.5H}{6.8M}$$

Osman is 87 kg and has consumed 3 standard drinks in the past hour. Estimate Osman's BAC to 3 decimak places using the following formula where $N$ represents the number of standard drinks, $H$ is the number of hours drinking and $M$ is his mass in kilograms.

$$BAC_{\text{Male}}=\frac{10N-7.5H}{6.8M}$$

$\begin{aligned}BAC_{\text{Male}}&=\frac{10N-7.5H}{6.8M}\\&=\frac{10(3)-7.5(1)}{6.8(87)}\\&=0.038032...\\&\approx0.038\end{aligned}$

### Reaching zero BAC

The number of hours required for a person to reach zero BAC after they stop consuming alcohol is given by the following formula:

::: {.formula}

$$\text{Time (h)}=\frac{BAC}{0.015}$$

This formula is **not** on the reference sheet.

You are **not** expected to remember this formula.
:::

Fiona has a BAC of 0.027. Use the following formula to estimate the time she will need for her BAC to reach zero. Answer to the nearest minute.

$$\text{Time}=\frac{BAC}{0.015}$$

Fiona has a BAC of 0.027. Use the following formula to estimate the time she will need for her BAC to reach zero. Answer to the nearest minute.

$$\text{Time}=\frac{BAC}{0.015}$$

<div class="row">
  <div class="column">
  $\begin{aligned}\text{Time}&=\frac{BAC}{0.015}\\&=\frac{(0.027)}{0.015}\\&=1.8\\&\quad(*)\\&=1^\circ48'0''\end{aligned}$

  Fiona needs to wait 1 hour and 48 minutes.
  </div>
  <div class="column">
  $(*)=$ <span style="font-family: casio">x</span>
  </div>
</div>

# Medication [![](pres.svg){width=16}](https://claytonpollard.github.io/stage6_standard_presentations/MS-A1/g_medication.pdf){target="blank"} {.tabset .tabset-fade .tabset-pills}

* calculate required medication dosages for children and adults from packets, given age or weight, using Fried’s, Young’s or Clark’s formula as appropriate
  + Fried's formula: Dosage for children 1-2 years $=\frac{\text{age (in months)}\ \times\ \text{adult dosage}}{150}$
  + Young's formula: Dosage for children 1-12 years $=\frac{\text{age of child (in years)}\ \times\ \text{adult dosage}}{\text{age of child (in years)}\ +\ 150}$
  + Clark's formula: Dosage $=\frac{\text{weight in kg}\ \times\ \text{adult dosage}}{70}$

### Converting concentrations

A concentration is a rate comparing a mass (g, mg, etc.) with a volume (L, mL, etc.). We need to consider both when converting a concentration from one unit to another.

The concentration of a mild analgesic is given as 80 mg per 50 mL. What is in g/mL?

The concentration of a mild analgesic is given as 80 mg per 50 mL. What is in g/mL?

$\begin{aligned}80\text{ mg}/50\text{ mL}&=\frac{80\text{ mg}}{50\text{ mL}}\\&=1.6\text{ mg/mL}\\&=1.6\div1000\text{ g/mL}\\&=0.0016\text{ g/mL}\end{aligned}$

### Calculating dosages

Often when calculating dosages you will be prescribed an amount in milligrams (mg) and need to calculate how much of a liquid that need to be taken. In these cases you can use the following:

::: {.formula}

$$\text{volume required}=\frac{\text{strength required}}{\text{strength of stock}}\times\text{volume of stock}$$

Where your *stock* is the medication you're given.

This formula is **not** on the reference sheet.

A patient is prescribed 1000 mg of a mild painkiller. The medication available contains 100 mg in 5 mL. How much medication should be given to the patient?

A patient is prescribed 1000 mg of a mild painkiller. The medication available contains 100 mg in 5 mL. How much medication should be given to the patient?

$\begin{aligned}\text{volume required}&=\frac{\text{strength required}}{\text{strength of stock}}\times\text{volume of stock}\\&=\frac{(1000\text{ mg})}{(100\text{ mg})}\times(5\text{ mL})\\&=50\text{ mL}\end{aligned}$

### Children and infants

There are three formulae used for children and infants:

::: {.formula}

* Fried's formula: Dosage for children 1-2 years $=\dfrac{\text{age (months)}\ \times\ \text{adult dosage}}{150}$
* Young's formula: Dosage for children 1-12 years $=\dfrac{\text{age of child (years)}\ \times\ \text{adult dosage}}{\text{age of child (years)}\ +\ 150}$
* Clark's formula: Dosage $=\dfrac{\text{weight (kg)}\ \times\ \text{adult dosage}}{70}$

These formulae are **not** on the reference sheet.

You are **not** expected to remember these formulae.

Jessica is 6 months old. Use Fried's formula to find the required infant dose if the adult dost is 20 mL.

$$\text{Dosage}=\dfrac{\text{age (months)}\ \times\ \text{adult dosage}}{150}$$

Jessica is 6 months old. Use Fried's formula to find the required infant dose if the adult dost is 20 mL.

$$\text{Dosage}=\dfrac{\text{age (months)}\ \times\ \text{adult dosage}}{150}$$

$\begin{aligned}\text{Dosage}&=\frac{\text{age (months)}\ \times\ \text{adult dosage}}{150}\\&=\frac{(6)\times(20)}{(150)}\\&=\frac{120}{150}\\&=0.8\text{ mL}\end{aligned}$
